---
title: Working With Your Fitbit Data in R
author: Matt Kaye
date: '2021-06-08'
categories: [R, data science]
lastmod: '2021-06-08T21:26:23-04:00'
toc: true
freeze: auto
---

```{r, include = F, echo = F}
knitr::opts_chunk$set(echo = T, message = F, warning = F, fig.path = "static/rmarkdown-libs", fig.align = 'center')
```

### Note: This post was updated as of 3/25/2023 for `fitbitr v0.3.0`

## Introduction

`fitbitr 0.1.0` is now available on CRAN! You can install it with

``` r
install.packages("fitbitr")
```

or you can get the latest dev version with

``` r
## install.packages("devtools")
devtools::install_github("mrkaye97/fitbitr")
```

`fitbitr` makes it easy to pull your Fitbit data into R and use it for whatever interests you: personal projects, visualization, medical purposes, etc.

This post shows how you might use `fitbitr` to pull and visualize some of your data.

## Sleep

First, you should either generate a new token with `generate_token()` or load a cached token with `load_cached_token()`.

```{r, include = F}
library(fitbitr)

generate_fitbitr_token()

head <- function(data, ...) {
  data |>
    utils::head(...) %>%
    knitr::kable("html") |>
    kableExtra::kable_styling(position = "center", full_width = TRUE)
}

```

```{r}
library(fitbitr)
library(lubridate)
library(tidyverse)

## Dates to use throughout post
start <- as_date("2020-01-01")
end <- as_date("2021-10-18")

generate_fitbitr_token()
```

And then you can start pulling your data!

```{r, cache = T}
sleep <- get_sleep_summary(
  start_date = end - months(3),
  end_date = end
)

head(sleep)
```

Once you've loaded some data, you can visualize it!

```{r}
library(zoo)
library(scales)
library(ggthemes)

sleep <- sleep %>%
  mutate(
   date = as_date(date),
   start_time = as_datetime(start_time),
   end_time = as_datetime(end_time),
   sh = ifelse(hour(start_time) < 8, hour(start_time) + 24, hour(start_time)), #create numeric times
   sm = minute(start_time),
   st = sh + sm/60,
   eh = hour(end_time),
   em = minute(end_time),
   et = eh + em/60,
   mst = rollmean(st, 7, fill = NA), #create moving averages
   met = rollmean(et, 7, fill = NA),
   year = year(start_time)
)

sleep %>%
    ggplot(aes(x = date)) +
    geom_line(aes(y = et), color = 'coral', alpha = .3, na.rm = T) +
    geom_line(aes(y = st), color = 'dodgerblue', alpha = .3, na.rm = T) +
    geom_line(aes(y = met), color = 'coral', na.rm = T) +
    geom_line(aes(y = mst), color = 'dodgerblue', na.rm = T) +
    scale_y_continuous(
      breaks = seq(0, 30, 2),
      labels = trans_format(
        function(x) ifelse(x > 23, x - 24, x), 
        format = scales::comma_format(suffix = ":00", accuracy = 1)
      )
    ) +
    labs(x = "Date", y = 'Time') +
    theme_fivethirtyeight() +
    scale_x_date(date_breaks = '1 month', date_labels = '%b', expand = c(0, 0)) +
    facet_grid(. ~ year, space = 'free', scales = 'free_x', switch = 'x') +
    theme(panel.spacing.x = unit(0,"line"), strip.placement = "outside")
```

This bit of code makes a nicely formatted plot of the times you went to sleep and woke up over the past three months. You can also use `fitbitr` to expand the time window with a little help from `purrr` (the Fitbit API rate limits you, so you can't request data for infinitely long windows in a single request).

```{r, cache = T}
## Pull three months of data
sleep <- map_dfr(
  3:0,
  ~ sleep_summary(
    end - months(.x), 
    end - months(.x) + months(1)
  )
)
```

After pulling the data, we can use the same code again to visualize it.

```{r}
sleep <- sleep %>%
  mutate(
   date = as_date(date),
   start_time = as_datetime(start_time),
   end_time = as_datetime(end_time),
   sh = ifelse(hour(start_time) < 8, hour(start_time) + 24, hour(start_time)), #create numeric times
   sm = minute(start_time),
   st = sh + sm/60,
   eh = hour(end_time),
   em = minute(end_time),
   et = eh + em/60,
   mst = rollmean(st, 7, fill = NA), #create moving averages
   met = rollmean(et, 7, fill = NA),
   year = year(start_time)
) %>%
  distinct()

sleep %>%
    ggplot(aes(x = date)) +
    geom_line(aes(y = et), color = 'coral', alpha = .3, na.rm = T) +
    geom_line(aes(y = st), color = 'dodgerblue', alpha = .3, na.rm = T) +
    geom_line(aes(y = met), color = 'coral', na.rm = T) +
    geom_line(aes(y = mst), color = 'dodgerblue', na.rm = T) +
    scale_y_continuous(
      breaks = seq(0, 30, 2),
      labels = trans_format(
        function(x) ifelse(x > 23, x - 24, x), 
        format = scales::comma_format(suffix = ":00", accuracy = 1)
      )
    ) +
    labs(x = "Date", y = 'Time') +
    theme_fivethirtyeight() +
  scale_x_date(date_breaks = '1 month', date_labels = '%b', expand = c(0, 0)) +
  facet_grid(. ~ year, space = 'free', scales = 'free_x', switch = 'x') +
  theme(panel.spacing.x = unit(0,"line"), strip.placement = "outside")
```

## Heart Rate and Steps

You can also pull your heart rate data with `fitbitr`. Maybe we're curious about seeing how the number of minutes spent in the "fat burn," "cardio," and "peak" zones correlates with the number of steps taken that day. Let's find out!

```{r, cache = T}
hr <- map_dfr(
  3:0,
  ~ heart_rate_zones(
    end - months(.x), 
    end - months(.x) + months(1)
  )
)

steps <- map_dfr(
  3:0,
  ~ steps(
    end - months(.x), 
    end - months(.x) + months(1)
  )
)
```

First, we can examine the heart rate data:

```{r}
head(hr)
```

and the steps data:

```{r}
head(steps)
```

Now, let's plot them against each other.

```{r}
df <- hr %>%
  filter(zone != "Out of Range") %>%
  group_by(date) %>%
  summarize(total_minutes = sum(minutes_in_zone), .groups = "drop") %>%
  inner_join(steps, by = "date")
  
df %>%
  mutate(steps = as.numeric(steps)) %>%
  filter(log(total_minutes) > 1) %>%
  ggplot(
    aes(
      steps,
      total_minutes
    )
  ) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  scale_x_log10() +
  scale_y_log10()
```

Or maybe it'd be interesting to predict your zone minutes from your steps:

```{r}
predictions <- df %>%
  mutate(steps = as.numeric(steps)) %>%
  lm(total_minutes ~ steps, data = .) %>%
  broom::tidy() %>%
  mutate(across(where(is.numeric), round, 5))

head(predictions)
```

## Wrapping Up

And that's it! Hopefully this helped show how `fitbitr` makes pulling your data easy, and gets you curious about the insights you can glean from your own data. The Fitbit API gives you access to so much interesting information about yourself, your habits, your fitness, and so much more, and `fitbitr` is just meant to be a door into that gold mine.
