% !TEX root = Matthew_Kaye_Resume.tex

\PassOptionsToPackage{dvipsnames}{xcolor}

\documentclass[10pt,a4paper,ragged2e,withhyper]{/Users/matt/documents/GitHub/mrkaye97.github.io/resume/altacv}

\geometry{left=1cm,right=1cm,top=1cm,bottom=1cm,columnsep=1cm}

\usepackage{paracol}

\usepackage[usenames,dvipsnames]{xcolor}

\usepackage[
	colorlinks = true,
	linkcolor = OrangeRed,
	urlcolor  = OrangeRed,
	citecolor = OrangeRed,
	anchorcolor = OrangeRed
]{hyperref}

\newcommand{\MYhref}[3][red]{\href{#2}{\color{#1}{#3}}}%

\ifxetexorluatex
\setmainfont{Roboto Slab}
\setsansfont{Lato}
\renewcommand{\familydefault}{\sfdefault}
\else
\usepackage[rm]{roboto}
\usepackage[defaultsans]{lato}
\renewcommand{\familydefault}{\sfdefault}
\fi

\definecolor{SlateGrey}{HTML}{2E2E2E}
\definecolor{LightGrey}{HTML}{666666}
\definecolor{DarkPastelRed}{HTML}{989898}
\definecolor{PastelRed}{HTML}{FF4D4D}
\definecolor{GoldenEarth}{HTML}{A9A9A9}
\colorlet{name}{black}
\colorlet{tagline}{PastelRed}
%\colorlet{heading}{DarkPastelRed}
\colorlet{headingrule}{GoldenEarth}
\colorlet{subheading}{PastelRed}
\colorlet{accent}{PastelRed}
\colorlet{emphasis}{SlateGrey}
\colorlet{body}{LightGrey}

\renewcommand{\namefont}{\Huge\rmfamily\bfseries}
\renewcommand{\personalinfofont}{\footnotesize}
\renewcommand{\cvsectionfont}{\LARGE\rmfamily\bfseries}
\renewcommand{\cvsubsectionfont}{\large\bfseries}

\renewcommand{\itemmarker}{{\small\textbullet}}
\renewcommand{\ratingmarker}{\faCircle}


\begin{document}
	\name{Matt Kaye}
	
	\personalinfo{%
		\homepage{mrkaye97.github.io}
		\github{mrkaye97}
		\email{mrkaye97@gmail.com}
		\phone{(646) 853-5997}
	}


	 \makecvheader
		
		\cvsection{Experience}
	
		\cvevent{Data Scientist}{CollegeVine}{Sept 2020 -- Present}{}
		\begin{itemize}
		\item First member of CollegeVine's data team. Responsible for all kinds of data projects, maintaining the codebase, integrating new tooling, and onboarding new team members. 
		\item Promoted a culture of shipping data scientific products with software engineering best practices, including unit and integration testing and a CI/CD process.
		\item Devised and implemented CollegeVine's Chancing model, used by hundreds of thousands of students and families every year to understand their chances of admission at their favorite schools.
		\item Created a recommender system to suggest new colleges to students. School recommendations also powers affinity scoring, a major component of our college product and sales process driving hundreds of thousands of dollars in revenue.
		\item Built a sequential testing toolkit to enable our team to call A/B tests more quickly without sacrificing statistical rigor, often cutting A/B test runtimes by half or more.
		\item Built out our first data warehouse for analytics, helping teammates uncover insights into all aspects of our business more quickly and correctly.
		\item Build and optimized data pipelines moving billions of rows from raw sources into our data warehouse to be consumed by downstream users.
		\item Introduced new tooling (dbt, Airflow, MLFlow, AWS stack) and owned its adoption by the team and integration into the codebase.
		\item Owned the deployment and monitoring of machine learning models in production, generally as REST APIs packaged up as Dockerized microservices and deployed on Heroku.
		\end{itemize}
		
		\divider
		

		\cvevent{Baseball Operations Fellow}{Baltimore Orioles}{Mar 2020 -- Sept 2020}{}
		\begin{itemize}
		\item Created a fully Bayesian, simulation-based projection system for MLB player performance over a six year time horizon
		\item Modeled free agent salaries with a gamma hurdle regression framework
		\item Devised a Markov Chain Monte Carlo approach to determining optimal shifts against opposing hitters
		\item Worked on a variety of day-to-day data science tasks related to game strategy and player evaluation
		\end{itemize}
		
		\divider 
				
		\cvsection{Projects \& Open-Source}

		\begin{itemize}
		\item Contributor to \MYhref{https://github.com/mlflow/mlflow}{MLFlow}, an open-source platform for managing the machine learning lifecycle.
		\item Current author and maintainer of \MYhref{https://github.com/mrkaye97/slackr}{slackr}, an R package for connecting R to Slack with 350k+ downloads.
		\item Contributor to \MYhref{https://github.com/statisticianinstilettos/recmetrics}{recmetrics}, a python library for evaluating recommender systems.
		\item Author and creator of \MYhref{https://github.com/collegevine/lightMLFlow}{lightMLFlow}, a lightweight, opinionated R wrapper for the MLFlow REST API.
		\item Author and creator of \MYhref{https://github.com/mrkaye97/fitbitr}{fitbitr}, an opinionated R wrapper for the Fitbit API.
		\end{itemize}

		\setcolumnwidth{4cm, 5cm,5cm} 
		\begin{paracol}{3}

		\cvsection{Education}
		
		\cvevent{Bachelor of Arts\\ Economics,  Mathematics}{Carleton College}{Sept 2016 -- Nov 2019}{}
		\divider\\
		\cvevent{}{Choate Rosemary Hall}{Sept 2013 -- June 2016}{}


		\switchcolumn
		
		\cvsection{Skills}
		\textbf{Programming:}\\
		\medskip
		\cvtag{Bash}
		\cvtag{Python}
		\cvtag{R}
		\cvtag{SQL}

		\smallskip
		\textbf{Frameworks, Software, and Tools:}\\
		\medskip

		\cvtag{AWS Tooling (Redshift, S3, Batch, etc.)} \\
		\cvtag{Airflow}
		\cvtag{CI/CD (Circle, GHA)}  \\
		\cvtag{dbt}
		\cvtag{Docker}
		\cvtag{Git}
		\cvtag{Heroku} \\
		\cvtag{MLFlow}
		\cvtag{\{\{plumber\}\}}
		\cvtag{\{\{shiny\}\}}

		\switchcolumn
		
		\cvsection{Hobbies}
		
		\cvtag{Distance running}\\
		\cvtag{Learning Haskell}\\
		\cvtag{Low and slow cooking}\\
		\cvtag{Nature \& architecture photography}	
		\cvtag{Reading fantasy novels}\\
		\cvtag{Skiing chop and powder}\\
		\cvtag{Solo traveling + hostel hopping}\\
		
		\end{paracol}
\end{document}
